---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r, echo=FALSE}
setwd("C:/Users/Nilmini/Dropbox/PhD/Coursework/JH Datascience/github/Course 5/Assignment 1/RepData_PeerAssessment1")
```

### 1. Code for reading in the dataset and/or processing the data
Unzip file, read and convert date variable into a date format.
```{r read}
unzip("activity.zip")
activity <- read.csv("activity.csv")
activity$date <- as.Date(activity$date)
str(activity)
head(activity)
```

### 2. Histogram of the total number of steps taken each day
Use ddply from the plyr package to create a dataframe holding the number of steps taken per day.
```{r histogram2-1}
library(plyr)
histodata <- ddply(activity,.(date),summarise,stepspd=sum(steps))
```

Produce the histogram
```{r histogram2-2}
hist(histodata$stepspd, ylim=range(0,10),breaks=25,xlim=range(0,25000),
     xlab="Number of steps per day",
     main="Frequency of number of steps taken per day from 1 Oct - 30 Nov 2012")
```


### 3. Mean and median number of steps taken each day
```{r meanmedian}
meand <- mean(histodata$stepspd, na.rm = TRUE)
mediand <- median(histodata$stepspd, na.rm = TRUE)
meand
mediand
```

The mean number of steps taken per day is `r as.integer(meand)` and the median number of steps taken each day is `r mediand`.


### 4. Time series plot of the average number of steps taken
Calculate the average steps by interval
```{r intervaldata}
intervaldata <- ddply(activity,.(interval),summarise,
                      stepspi=mean(steps, na.rm = TRUE))
```

Plot average steps by interval
```{r intervalplot}

plot(intervaldata$interval, intervaldata$stepspi, type="l",
     main="Average number of steps taken in each time interval in Oct-Nov 2012",
     xlab="5 minute time interval",
     ylab="Average number of steps in time interval",
     xlim=range(0,2500))
```



### 5. The 5-minute interval that, on average, contains the maximum number of steps
```{r fivemax}
row <- which.max(intervaldata$stepspi)
x <- intervaldata[row,1]
x
```

The 5 minute interval labeled as `r x` has the highest average number of steps taken in the October to November time period.


### 6. Code to describe and show a strategy for imputing missing data

First, look for patterns in the missing data.

1. check which days are missing data, and how much data is missing
```{r missingdata1}
#create variables storing number of days from first data point
startdate <- activity[1,2]
activity$numdays <- difftime(activity$date,startdate,units="days")

naperday <- table(activity$steps, activity$numdays,useNA="always")
naperday <- tail(naperday,1)

barplot(naperday,
        ylim=range(0,300), las=2, cex.axis = 0.5, cex.names=0.5,
        xlab = "Day", ylab="Number of missing datapoints",
        main="Missing data points by day")
```

Data are missing for entire days at a time, i.e. 288 missing data points.
There does not appear to be a pattern over the 60 days of when there a days missing data.


Check whether there are certain days of the week that have missing data
```{r}
activity$days <- weekdays(activity$date)
nadayofweek <- table(activity$steps, activity$days,useNA="always")
nadayofweek <- tail(nadayofweek,1)
nadayofweek
```
There does not appear to be a strong pattern for which days of the week are missing data.

Next, look for patterns in the data that will be used to impute.
Calculate the average steps by interval and day.
```{r intervaldaydata}
intervaldataday <- ddply(activity,.(interval, days),summarise,
                      stepspid=mean(steps, na.rm = TRUE))
```

Plot average steps by interval
```{r intervaldayplot}
library(ggplot2)
p <- qplot(interval,stepspid,data=intervaldataday, 
        geom="line",
        xlim=range(0,2500),
        ylab="Average number of steps in time interval",
        xlab="5 minute time interval and day of week",
        main="Average number of steps taken in each time interval by day in Oct-Nov 2012")
p <- p + facet_wrap(~days, ncol=3)
p
```

It appears that the pattern of activity changes for each day of the week, as well as by 5 minute interval.

Therefore, we should impute missing data using the mean number of steps by interval and day.


```{r impute missing data}
# Extract complete data
missing <- activity[is.na(activity$steps),]
complete <- activity[!is.na(activity$steps),]

missing <- merge(missing, intervaldataday, by=c("days", "interval"))
missing$steps <- missing$stepspid
missing <-subset(missing, select= -c(stepspid))

imputed <- rbind(missing,complete)
imputed <- subset(imputed, select=-c(numdays, days))
imputed <- imputed[order(imputed$date,imputed$interval),]
imputed$id <- seq.int(nrow(imputed))
rownames(imputed) <- imputed$id
imputed <- subset(imputed,select=-c(id))
imputed <- imputed[,c("steps", "date", "interval")]
str(imputed)
```

The complete data looks like this:
```{r}
head(imputed)
```

Compared to the original dataset, which looks like this:
```{r}
head(activity)
```

###7. Histogram of the total number of steps taken each day after missing values are imputed

Use ddply from the plyr package to create a dataframe holding the number of steps taken per day, and produce histogram
```{r histogram71}
histodata2 <- ddply(imputed,.(date),summarise,stepspd=sum(steps))

hist(histodata2$stepspd, ylim=range(0,10),breaks=25,xlim=range(0,25000),
     xlab="Number of steps per day",
     main="Frequency of number of steps taken per day 
     from 1 Oct - 30 Nov 2012 (missing data imputed)")
```

For comparison, this histogram (left) is displayed next to the previous histogram (right), and it can be seen that they look very similar, with some changes in frequency around the centre of the graph.

```{r oldhistogram}
par(mfrow=c(1,2))
hist(histodata2$stepspd, ylim=range(0,10),breaks=25,xlim=range(0,25000),
     xlab="Number of steps per day",
     main="Frequency of number of steps taken per day 
     from 1 Oct - 30 Nov 2012 (missing data imputed)", cex.main=0.7)

hist(histodata$stepspd, ylim=range(0,10),breaks=25,xlim=range(0,25000),
     xlab="Number of steps per day",
     main="Frequency of number of steps taken per day 
     from 1 Oct - 30 Nov 2012 (no imputation)", cex.main=0.7)

```


### Mean and median number of steps taken each day
```{r meanmedian7}
meand2 <- mean(histodata2$stepspd, na.rm = TRUE)
mediand2 <- median(histodata2$stepspd, na.rm = TRUE)
meand2
mediand2
```

The mean number of steps taken per day is `r as.integer(meand2)` and the median number of steps taken each day is `r as.integer(mediand2)`. 
In the previous analysis, where missing data were dropped, the mean is similar (`r as.integer(meand)`) and the median is slightly less (`r as.integer(mediand)`).

This suggests that the imputation method of using mean values in the corresponding interval and day of the week, as one would expect, only changes the median and mean slightly.


### 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends


```{r}
imputed$days <- weekdays(imputed$date)
imputed$wwe <- ifelse((imputed$days=="Saturday"|imputed$days=="Sunday"), "WEEKEND", "WEEKDAY")
imputed$wwe <- factor(imputed$wwe)
table(imputed$days, imputed$wwe) ##check
```


Calculate the average steps by interval and day
```{r}
intervaldatawwe <- ddply(imputed,.(interval, wwe),summarise,
                      stepspid=mean(steps))

```

Plot average steps by interval
```{r}
library(ggplot2)
p <- qplot(interval,stepspid,data=intervaldatawwe, 
        geom="line",
        xlim=range(0,2500),
        ylab="Average number of steps in time interval",
        xlab="5 minute time interval",
        main="Comparison of average number of steps taken in each time interval on weekdays and weekends in Oct-Nov 2012")
p <- p + facet_wrap(~wwe, ncol=2)
p
```

The graphs show that after a rest period, steps are taken earlier on weekdays, around interval 500, with a large spike in the number of steps taken between interval 500 and 1000. On the weekends however, steps gradually increase from around interval 600, and fluctuate between intervals 750 and 2000 at slightly higher levels than on weekdays, although without a large spike early in the day as seen on weekdays. 